---
title: "Data Engineering: **Embracing** the UNIX philosophy - Part 2"
date: "2023-07-12"
---

```{r slides-videos, echo=FALSE, include=FALSE}
source(here::here("R", "slide-things.R"))
```

## Slides

The slides for today's lesson are available online as an PDF file. Use the 
button below to open the slides either in a new tab in your browser. Or you can
just scroll the embedded slides below. For the embedded slides below, you can 
also click in the slides and navigate through them with your left and right 
arrow keys, for convenience.

```{r show-slide-tabs, echo=FALSE, results="asis"}
slide_buttons("/data_eng/_files/04-slides-data-eng", output_type = "pdf")
```
{{\< downloadthis ./_files/raw_data/ninja-way.csv
dname="ninja-way" label="Download ninja-way.csv" icon="file-earmark"
type="success" \>}}

```{=html}
<iframe width="780" height="500" src="/data_eng/_files/04-slides-data-eng.pdf" title="Lecture 4"></iframe>
```

## Creating the `ninja-way.csv` file manually

You can use the following instructions at the bash command line in your desired
working directory to create, i.e., reproduce the `ninja-way.csv` file manually.

::: {.callout-note collapse="true"}

## Instructions to create the `ninja-way.csv` file manually

In order to create the `ninja-way.csv` file as used in these lecture notes, you
will need to perform the following steps in a bash terminal session:

1.  First run `cat > ninja-way.csv`
2.  The prompt will move to the next line. So `cat` is waiting for **input**
    from you from `stdin`, i.e., your keyboard!
3.  Copy paste the following text into the terminal, where the prompt is active.

```{txt}
This is a nice csv containing characters from the Anime: Naruto
This is based on a manga by various authors
See the following fields which contain the data
id,first_name,last_name,village,season_first_appearance,home
1,Naruto,Uzamaki,leaves,1,leaves village
1,Naruto,Uzamaki,leaves,1,leaves village
1,Naruto,Uzamaki,leaves,1,leaves village
2,Sasuke,Uchiha,leaves,1,leaves village


3,Sakura,Haruno,leaves,1,leaves village
TODO: add more leaves village characters

4,Gaara,None,sand,2,sand village
4,Gaara,None,sand,2,sand village
5,Temari,Nara,sand,2,sand village



## we should add more sand village characters

6,Sai,Yamanaka,leaves,4,leaves village

#closing the file now

```
4.  Make sure that you hit the `Return` key after the last pasted line. This
    will insert a blank line after the pasting is complete. This is important to
    do else you will end up omitting the final commented line in the above
    pasted text.
5.  You can now hit `Ctrl-C` to terminate the `cat` input command. The
    `ninja-way.csv` file should now be created.
6.  Run `less ninja-way.csv` to view the newly created file and check that you
    get the same contents as those that you just pasted in Step 3.

:::
