---
title: "Schedule"
sidebar: false
toc: false
page-layout: full
---

```{r packages-data, include=FALSE}
library(tidyverse)
library(knitr)
library(kableExtra)

withr::with_dir(here::here(), {
  schedule_nested <- targets::tar_read(schedule_page_data)
  targets::tar_load(schedule_ical_file)
})

base_url <- yaml::read_yaml(here::here("_quarto.yml"))$website$`site-url`
ical_url_absolute <- paste0(base_url, "/", schedule_ical_file)
```

::::: {.schedule-page}

Here's your roadmap for the semester!

  - [**Lecture**](/class/index.qmd) ({{\< fa book-open-reader \>}}): This page
    contains the slides (and additional readings) for the data science or data
    engineering lectures. These materials will typically be released *right*
    after each class, and in some cases, *directly before* the class.

  - [**Assignment**](/assignment/index.qmd) ({{\< fa pen-ruler \>}}): This page
    contains the instructions for each assignment, i.e., demo-lab or final
    research project materials.

```{r build-table, include=FALSE}
show_table <- function(group_id) {
  # Add a heading
  cat(as.character(paste("\n\n###", schedule_nested$group[[group_id]], "\n\n")))
  
  # Make the table
  tbl <- schedule_nested$data[[group_id]] %>% 
    select(-subgroup) %>% 
    kbl(escape = FALSE, align = "rlcc", table.attr = 'class="schedule-table"') %>% 
    kable_styling() %>% 
    column_spec(1, width = "15%", extra_css = "padding-right: 20px;") %>% 
    column_spec(2, width = "55%") %>% 
    column_spec(3:4, width = "10%") %>% 
    pack_rows(index = schedule_nested$subgroup_index[[group_id]],
              label_row_css = "border-bottom: 1px solid #adb5bd;")
  
  cat(tbl)
}
```

```{r show-table, echo=FALSE, results="asis"}
walk(seq(1, nrow(schedule_nested)), ~show_table(.x))
```

:::::
