---
title: "Introduction and Welcome to SURE 2023"
date: "2023-06-05"
---

```{r slides-videos, echo=FALSE, include=FALSE}
source(here::here("R", "slide-things.R"))
```

<!-- ## Readings -->

<!-- - The [syllabus](/syllabus.qmd), [class](/class/index.qmd), and [assignments](/assignment/index.qmd) pages for this class -->

<!-- - {{< fa file-pdf >}}&ensp;[@Nelson:2021] -->

<!-- - {{< fa file-pdf >}}&ensp;The "How to Read" section in [@Green:2021] -->

<!-- - {{< fa file-pdf >}}&ensp;[@Powner:2016] -->

<!-- - {{< fa file-pdf >}}&ensp;[@MeierHill:2005] -->


## Slides

The slides for today's lesson are available online as an PDF file. Use the 
buttons below to open the slides either as an interactive website. You can also 
click in the slides below and navigate through them with your left and right 
arrow keys.

```{r show-slide-tabs, echo=FALSE, results="asis"}
slide_buttons("/slides/01-slides", output_type = "pdf")
# slide_tabs(slide_details, "/slides/01-slides.html")
```

```{=html}
<iframe width="780" height="500" src="/slides/01-slides.pdf" title="Lecture 1"></iframe>
```

```{=html}
<iframe width="780" height="500" src="/slides/02-slides.pdf" title="Lecture 2"></iframe>
```

```{=html}
<iframe width="780" height="500" src="/slides/03-slides.pdf" title="Lecture 3"></iframe>
```

```{=html}
<iframe width="780" height="500" src="/slides/04-slides.pdf" title="Lecture 4"></iframe>
```

```{=html}
<iframe width="780" height="500" src="/slides/05-slides.pdf" title="Lecture 5"></iframe>
```

```{=html}
<iframe width="780" height="500" src="/slides/06-slides.pdf" title="Lecture 6"></iframe>
```

```{=html}
<iframe width="780" height="500" src="/slides/07-slides.pdf" title="Lecture 7"></iframe>
```

```{=html}
<iframe width="780" height="500" src="/slides/08-slides.pdf" title="Lecture 8"></iframe>
```

```{=html}
<iframe width="780" height="500" src="/slides/09-slides.pdf" title="Lecture 9"></iframe>
```

```{=html}
<iframe width="780" height="500" src="/slides/10-slides.pdf" title="Lecture 10"></iframe>
```

```{=html}
<iframe width="780" height="500" src="/slides/11-slides.pdf" title="Lecture 11"></iframe>
```

```{=html}
<iframe width="780" height="500" src="/slides/12-slides.pdf" title="Lecture 12"></iframe>
```

```{=html}
<iframe width="780" height="500" src="/slides/13-slides.pdf" title="Lecture 13"></iframe>
```

```{=html}
<iframe width="780" height="500" src="/slides/14-slides.pdf" title="Lecture 14"></iframe>
```

```{=html}
<iframe width="780" height="500" src="/slides/15-slides.pdf" title="Lecture 15"></iframe>
```

```{=html}
<iframe width="780" height="500" src="/slides/16-slides.pdf" title="Lecture 16"></iframe>
```

```{=html}
<iframe width="780" height="500" src="/slides/17-slides.pdf" title="Lecture 17"></iframe>
```

```{=html}
<iframe width="780" height="500" src="/slides/18-slides.pdf" title="Lecture 18"></iframe>
```

```{=html}
<iframe width="780" height="500" src="/slides/19-slides.pdf" title="Lecture 19"></iframe>
```

```{=html}
<iframe width="780" height="500" src="/slides/20-slides.pdf" title="Lecture 20"></iframe>
```

```{=html}
<iframe width="780" height="500" src="/slides/21-slides.pdf" title="Lecture 21"></iframe>
```

```{=html}
<iframe width="780" height="500" src="/slides/22-slides.pdf" title="Lecture 22"></iframe>
```

```{=html}
<iframe width="780" height="500" src="/slides/23-slides.pdf" title="Lecture 23"></iframe>
```

```{=html}
<iframe width="780" height="500" src="/slides/24-slides.pdf" title="Lecture 24"></iframe>
```


## Plan for the day

- **Who are you all and why are you here?**

- **The pains and wonders of bureaucracies + building our Global Bureaucracy**^[Good luck spelling this word correctly over and over and over this semester!] **Toolkit^™^**
  - [Jan Banning's "Bureaucratics"](https://www.janbanning.com/gallery/bureaucratics/)

- **How the class works**
  - [Mini presentation signup sheet](https://docs.google.com/spreadsheets/d/1KZZ7Ei85cb97IjnTVwiYsUiSiOSgtbc_pHVuRxPWgz4/edit?usp=sharing)

- **Things they don’t teach you in grad school**

  - **How to read and annotate an article**

    - [Zotero](https://www.zotero.org/) + [Zotero Connector](https://www.zotero.org/download/connectors) + [GSU's Zotero guide](https://research.library.gsu.edu/zotero)
    - [Class Zotero group library](https://www.zotero.org/groups/compasp23)
    - [What are DOIs?](https://apastyle.apa.org/style-grammar-guidelines/references/dois-urls)
    - Taking and storing notes:
      - [Zotero](https://www.zotero.org/)
      - [Obsidian](https://obsidian.md/)
      - [Notion](https://www.notion.so/)
      - [OneNote](https://www.onenote.com/?public=1)
      - [Evernote](https://evernote.com/)
    - Things we'll read together:
      - {{< fa file-pdf >}}&ensp;[@Wilson:1887]
      - {{< fa file-pdf >}}&ensp;[@MilwardJensenRoberts:2016]
      - {{< fa file-pdf >}}&ensp;[@Roberts:2018]

  - [**Break!**]{.text-success}

  - **How to summarize, synthesize, and write about articles**

    - [Zotero in Word](https://www.zotero.org/support/word_processor_plugin_usage) + [Zotero in Google Docs](https://www.zotero.org/support/google_docs) + Zotero [in RStudio](https://posit.co/blog/rstudio-1-4-preview-citations/) or [with Markdown in general](https://www.andrewheiss.com/blog/2023/01/08/bibdesk-to-zotero-pandoc/#example-workflow-with-quarto-r-markdown-pandoc)
    - [Summary vs. synthesis](https://www.youtube.com/watch?v=R_rnkM7jvlY) + [The Synthesis Test](https://www.youtube.com/watch?v=R_rnkM7jvlY&t=241s)

  - **How to discuss an article in a seminar**

  	- General summary of the article
  	- Strengths
  	- Weaknesses
  	- How does it connect to other readings?
  	- How does it connect to the broader course themes?
  	- How does it connect to current events?
  	- What does it add to our Global Bureaucracy Toolkit^™^?
