---
title: "The grammar of graphics and ggplot2"
date: "2023-06-07"
---

```{r slides-videos, echo=FALSE, include=FALSE}
source(here::here("R", "slide-things.R"))
```

## Slides

The slides for today's lesson are available online as an PDF file. Use the 
button below to open the slides either in a new tab in your browser. Or you can
just scroll the embedded slides below. For the embedded slides below, you can 
also click in the slides and navigate through them with your left and right 
arrow keys, for convenience.

```{r show-slide-tabs, echo=FALSE, results="asis"}
slide_buttons("/slides/03-slides", output_type = "pdf")
```
{{\< downloadthis ../slides/03-slides.Rmd
dname="03-slides" label="Download Rmd of all slides" icon="file-earmark"
type="success" \>}}

```{=html}
<iframe width="780" height="500" src="/slides/03-slides.pdf" title="Lecture 1"></iframe>
```

These were additional supplementary slides which [Shamindra](/people/staff/shrotriya-shamindra.qmd)
went through to describe the `%>%` operator and function composition.

```{=html}
<iframe width="780" height="500" src="/class/_files/data_sci/03-slides-supp.pdf" title="Lecture 1"></iframe>
```
